---
title: "Mueller_Rgr_2023"
output: html_notebook
---



### Load libraries
```{r, message=FALSE}
  
library(Seurat)
library(tidyverse)
library(cowplot)
library(gridExtra)
  
```
  
### Load datasets
```{r}
  
##########
# Mouse scRNA
seu.mouse.Campello <- readRDS("GSE230049_Campello_2023.rds")
seu.mueller <- readRDS("../data/interim/seu.meuller.rds")

##########
# Human scRNA
seu.human.Yan <- readRDS("GSE148077_Yan_2020.rds")
fov_per <- ifelse(seu.human.Yan$Source == "Fovea", "Fovea", "Periphery")
seu.human.Yan <- AddMetaData(seu.human.Yan, metadata = fov_per, col.name = "Location")

##########
# Macaque scRNA
seu.macaque.Peng <- readRDS("GSE118546_Peng_2019.rds")
  
```
  
### Create directories
```{r}
  
# expt_nm <- "RGR_Mueller_glia"
dir_export <- file.path("~")
  
```
  
### Mouse Rgr positive Mueller glia cells
```{r}
  
# Subset Mueller glia
mueller_cells <- seu.mouse.Campello$CellType == "Mueller cell"

# Calculate number of Rgr positive cells
rgr_cells <- sum(GetAssayData(seu.mouse.Campello, 
                              assay = "decontXcounts", 
                              slot = "counts")["Rgr", mueller_cells]>0)

##########
# Percentage Rgr positive cells
mm_rgr_percent <- rgr_cells / length(mueller_cells[mueller_cells == T])
  
```
  
### Human RGR positive Mueller glia cells
```{r}
  
# Subset Mueller glia
seu.human.Yan.mg <- subset(seu.human.Yan, subset = CellType == "Muller")

# Set retina location information
mg_loc <- table(seu.human.Yan.mg$Location)
mg_fov <- seu.human.Yan.mg$Location == "Fovea"
mg_per <- seu.human.Yan.mg$Location == "Periphery"

# Calculate number of RGR positive cells
rgr_cells_fov <- sum(GetAssayData(seu.human.Yan.mg, 
                                  assay = "RNA", 
                                  slot = "counts")["RGR", mg_fov]>0)
rgr_cells_per <- sum(GetAssayData(seu.human.Yan.mg, 
                                  assay = "RNA", 
                                  slot = "counts")["RGR", mg_per]>0)

##########
# Percentage RGR positive cells
rgr_percent_hs <- round(c(rgr_cells_fov, rgr_cells_per) / mg_loc * 100, digits = 1)
  
##########
# Violin Plot of RGR expression in Mueller glia
v <- VlnPlot(seu.human.Yan.mg, 
             group.by = "Location", 
             features = "RGR", 
             assay = "RNA", 
             slot = "counts",
             pt.size = 0.01, ) +
  labs(caption = paste0("RGR positive Mueller cells (%)\nFovea: ", 
                        rgr_percent_hs["Fovea"], 
                        "\nPeriphery: ", 
                        rgr_percent_hs["Periphery"])) +
  xlab(NULL)

cowplot::ggsave2(plot = v, 
                 filename = file.path(dir_export, "Human-Yan_MG_Vln_RGR_region.pdf"), 
                 height = 6, width = 6)
  
```
  
### Macaque RGR positive Mueller glia cells
```{r}
  
# Subset Mueller glia
seu.macaque.Peng.mg <- subset(seu.macaque.Peng, subset = CellType == "MG")

# Set retina location information
mg_loc <- table(seu.macaque.Peng.mg$Subcluster)
mg_fov <- seu.macaque.Peng.mg$Subcluster == "Fovea"
mg_per <- seu.macaque.Peng.mg$Subcluster == "Periphery"

# Calculate number of RGR positive cells
rgr_cells_fov <- sum(GetAssayData(seu.macaque.Peng.mg, 
                                  assay = "RNA", 
                                  slot = "counts")["RGR", mg_fov]>0)
rgr_cells_per <- sum(GetAssayData(seu.macaque.Peng.mg, 
                                  assay = "RNA", 
                                  slot = "counts")["RGR", mg_per]>0)

##########
# Percentage RGR positive cells
rgr_percent_macaque <- round(c(rgr_cells_fov, rgr_cells_per) / mg_loc * 100, digits = 1)
  
##########
# Violin Plot of RGR expression in Mueller glia
v <- VlnPlot(seu.macaque.Peng.mg, 
             group.by = "Subcluster", 
             features = "RGR", 
             assay = "RNA", 
             slot = "counts",
             pt.size = 0.01, ) +
  labs(caption = paste0("RGR positive Mueller cells (%)\nFovea: ", 
                        rgr_percent_macaque["Fovea"], 
                        "\nPeriphery: ", 
                        rgr_percent_macaque["Periphery"])) +
  xlab(NULL)

cowplot::ggsave2(plot = v, 
                 filename = file.path(dir_export, "Macaque-Peng_MG_Vln_RGR_region.pdf"), 
                 height = 6, width = 6)
  
```

### Dotplots of visual cycle genes in Mouse, Human, and Macaque cell types
```{r}

# Visual cycle genes of interest
gois <- c("RGR", "RLBP1", "LRAT", "DHRS3", "RPE65", "RDH5", "RDH8", "RDH10", "RDH11", "RDH12")


####################
# Mouse scRNA data
####################

# Set cell order for x-axis in plot
seu.mouse.Campello$CellType <- factor(as.character(seu.mouse.Campello$CellType), levels = rev(as.character(Misc(seu.mouse.Campello, slot = "cell_order"))), ordered = T)

##########
# Make dotplot of expression for genes-of-interest
d <- DotPlot(seu.mouse.Campello, 
             group.by = "CellType", 
             features = str_to_title(gois), 
             assay = "decontXcounts") +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.95, hjust = 0.9)) +
  labs(title = "Mouse") +
  labs(x = NULL, y = NULL)

cowplot::ggsave2(plot = d, 
                 filename = file.path(dir_export, paste0("Mouse-Campello_Dot_gois.pdf")), 
                 height = 4, width = 6)


####################
## Human scRNA data
####################

# Subset data via retina location for preparation of plotting
seu.human.Yan$CellType <- factor(seu.human.Yan$CellType, levels = rev(levels(Idents(seu.human.Yan))))
seu.human.fov <- subset(seu.human.Yan, subset = Location == "Fovea")
seu.human.per <- subset(seu.human.Yan, subset = Location == "Periphery")

##########
# Make dotplot of expression for genes-of-interest in fovea and peripheral retina
d1 <- DotPlot(seu.human.fov, 
              group.by = "CellType", 
              features = gois, 
              scale = F) +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.95, hjust = 0.9)) +
  labs(title = "Human", subtitle = "Fovea") +
  labs(x = NULL, y = NULL)


d2 <- DotPlot(seu.human.per, 
              group.by = "CellType", 
              features = gois, 
              scale = F) +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.95, hjust = 0.9)) +
  labs(subtitle = "Periphery") +
  labs(x = NULL, y = NULL)

g <- gridExtra::grid.arrange(grobs=list(d1, d2), ncol=1)

cowplot::ggsave2(plot = g, 
                 filename = file.path(dir_export, paste0("Human-Yan_Dot_gois.pdf")), 
                 height = 8, width = 6)


####################
# Macaque scRNA data
####################

# Subset data via retina location for preparation of plotting
cell_order_macaque <- levels(Idents(seu.macaque.Peng))[c(12,11,10,4,3,5,2,1,13,6,9,8,7)]
seu.macaque.Peng$CellType <- factor(seu.macaque.Peng$CellType, levels = rev(cell_order_macaque))
seu.macaque.fov <- subset(seu.macaque.Peng, subset = Subcluster == "Fovea")
seu.macaque.per <- subset(seu.macaque.Peng, subset = Subcluster == "Periphery")

##########
# Make dotplot of expression for genes-of-interest in fovea and peripheral retina
d1 <- DotPlot(seu.macaque.fov, 
              group.by = "CellType", 
              features = gois, 
              scale = F) +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.95, hjust = 0.9)) +
  labs(title = "Macaque", subtitle = "Fovea") +
  labs(x = NULL, y = NULL)

d2 <- DotPlot(seu.macaque.per, 
              group.by = "CellType", 
              features = gois, 
              scale = F) +
  theme(axis.text.x = element_text(angle = 60, vjust = 0.95, hjust = 0.9)) +
  labs(subtitle = "Periphery") +
  labs(x = NULL, y = NULL)

g <- gridExtra::grid.arrange(grobs=list(d1, d2), ncol=1)

cowplot::ggsave2(plot = g, 
                 filename = file.path(dir_export, paste0("Macaque-Peng_Dot_gois.pdf")), 
                 height = 8, width = 6)
  
```



### Mouse Differential Gene Expression in Mueller glia cells, Rgr Pov vs Neg
```{r}

####################
# Prep Seurat object for DGE
####################

# Subset Mueller cells
seu.mueller <- subset(seu, idents = "Mueller cell")

# Remove one outlier cell
outlier <- "M24_whole_M24_whole_GGTGTCGTCTTACCAT-1"
good_cells <- Cells(seu.mueller)[!(Cells(seu.mueller) %in% outlier)]
seu.mueller <- subset(seu.mueller, cells = good_cells)

# Add meta column indicating positive expression of Rgr using SCT threshold
rgr_exp <- GetAssayData(seu.mueller, assay = "SCT", slot = "counts")["Rgr", ]>0
rgr_exp <- ifelse(rgr_exp == T, "rgr_pos", "rgr_neg")
seu.mueller <- AddMetaData(seu.mueller, rgr_exp, col.name = "rgr_exp_sct")


####################
# Perform DGE analysis
####################

# Parameters for DGE
pct <- 0.1
fc <- 1

# Perform DGE
seu.mueller <- SetIdent(seu.mueller, value = "rgr_exp_sct")
seu.mueller <- PrepSCTFindMarkers(seu.mueller)
rgr.de.sct <- FindMarkers(seu.mueller, 
                          assay = "SCT", 
                          slot = "data",
                          ident.1 = "rgr_pos", 
                          ident.2 = "rgr_neg",
                          logfc.threshold = log(fc),
                          recorrect_umi = F)

# Filter for genes with better than 1% FDR
rgr.de.sct <- rgr.de.sct %>% 
  arrange(p_val_adj, desc(abs(avg_log2FC))) %>% 
  filter(p_val_adj < 0.01)

# Export results table
write.table(rgr.de.sct, 
            file = file.path(dir_export, paste0("DGE_Mouse_Rgr_posVneg_SCTclassif.tsv")), 
            quote = F, sep = "\t")
  
```


### Macaque Differential Gene Expression in Mueller glia cells, Rgr Pov vs Neg
```{r}

####################
# Prep Seurat object for DGE
####################

# Add meta column indicating positive expression of Rgr
rgr_exp <- GetAssayData(seu.macaque.Peng.mg, assay = "RNA", slot = "counts")["RGR", ]>0
rgr_exp <- ifelse(rgr_exp == T, "rgr_pos", "rgr_neg")
seu.macaque.Peng.mg <- AddMetaData(seu.macaque.Peng.mg, rgr_exp, col.name = "rgr_exp")


####################
# Perform DGE analysis
####################

# Parameters for DGE
pct <- 0.1
fc <- 1

# Perform DGE
seu.macaque.Peng.mg <- SetIdent(seu.macaque.Peng.mg, value = "rgr_exp")
rgr.de.macaque <- FindMarkers(seu.macaque.Peng.mg, 
                              assay = "RNA", 
                              slot = "counts", 
                              ident.1 = "rgr_pos", 
                              ident.2 = "rgr_neg", 
                              logfc.threshold = log(fc),
                              recorrect_umi = F)

# Filter for genes with better than 1% FDR
rgr.de.macaque <- rgr.de.macaque %>% 
  arrange(p_val_adj, desc(abs(avg_log2FC))) %>% 
  filter(p_val_adj < 0.01)

# Export results table
write.table(rgr.de.macaque, 
            file = file.path(dir_export, paste0("DGE_Macaque_Rgr_posVneg_mueller.tsv")), 
            quote = F, sep = "\t", )
  
```

